What We Eval in the Shadows

### 2021-04-17 paper/main.tex
5.3

Contrary to the Javascript \eval, which does not precise the scope in which to evaluate,the second argument of \eval in R, \c{envir}, specifically is dedicated to that.
The environment in which evaluation happens determines what is visible to the
computation started by \eval and the potential reach of its side-effects. The
second argument to \eval specifies that environment. Environments of \eval can be
classified into the following four kinds:

@@

\noindent
As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
function environment with the \c{global} kind as a distant second. This means
that most variable lookups and most side effects either read/update existing
local variables or introduce new ones. But for which function? From the point of
view of the function that called \eval, Table~\ref{tab:funoffset} gives an
offset on the call stack. Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on. The data suggests that in 81\% of cases, \eval accesses its caller's environment -- this means  the variables of the function where \eval textually
occurs are read and written to. It is interesting that some 1.5\% of sites
evaluate a code in an environment that is three frames or above from \eval. This
distance implies that, in general, modular reasoning is impossible. Understanding the behavior of any piece of code entails fully understanding the
behavior of all functions that the code may call as the actions of \eval may
happen at a distance. Finally, note that any particular site may have several
kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the cases,  a site has a single kind. It means that a JIT compiler can predict the \emph{kind} of  expression passed to \eval in \packageNbOneCategoryEnvirSitePercent of the sites by just examining at the first invocation of a site.

@@

\noindent
Sites that evaluate in the global environment are likely split between
intentional and accidental use. Direct references to the top-level, using
\c{globalenv()} or \c{.GlobalEnv}, are rare; they occur in only
\packageNbExplicitGlobalSites sites. We suspect that the majority of uses of the
global environment come from the fact that our corpus consists of scripts loaded at the top level. Thus, that environment may often be the caller of
\eval or close to it. The reason we make this point is that values stored in the
global environment are visible to all functions and are not reclaimed by the
garbage collector. So accidental uses may pollute that namespace.

Synthetic environments need a parent. It is specified as an argument of the \c{new.env} function, or when \c{envir} is a list or a data frame, \eval uses its third argument
(\c{enclos}) as the parent.  The parent is used to lookup variables not found in its
child (side effects go to the child only). Table~\ref{tab:newenvs} shows the
kinds of the parents of synthetic environments. Most of them are functions, then
come global and package.

\paragraph{Discussion}
The data presented in this section is what one could expect. ({\it Function})
The majority of \eval call sites access the environment of some function on the
call stack. Usually, the current function, but sometimes frames arbitrarily far
up the call stack. One thing the data does not say is how that environment was
obtained. The expected case is that the environment was the one obtained from a
promise, as promises combine code and the environment in which that code
originates. Likely less frequent are cases where \eval is provided the results
of programmatically selecting some call stack. ({\it Synthetic}) The relatively
high frequency of synthetic environments corresponds to cases when one wants to
evaluate an expression in either a restricted environment, use a data
structure as an environment. ({\it Global}) The fact that the global environment
shows up so frequently is likely an artifact of how the code is run. ({\it
  Package}) There are only \packageNbPackageNamespaceEnvSites sites that use a %77 sites
 package environment in \eval. This is probably best as it is an improper form of adding bindings or modifying functions of a loaded package. Many more sites instead wrap a package environment in a new
environment to create singletons in a package.

%% The \c{imchange} function of package \c{imager} makes it possible to modify
%% images using a dedicated formula syntax.\footnote{Inspired by {map} in
%% package \emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a
%% new environment that inherits from \c{parent.frame()} by default (classified
%% as 1+).
%% \begin{lstlisting}
%%  newenv <- new.env()
%%  ...
%%  fo <- parse(text=as.character(fo)[2])
%%  im[where] <- eval(fo,envir=newenv,enclos=env)
%% \end{lstlisting}

%% \c{adjCoef} in package \emph{actuar} find the root of an equation defined by
%% a function \c{h} whose arguments must be named \c{x} and \c{y}. \c{h} is
%% transformed into an auxiliary function \c{h2} that can be optimized. Here,
%% the list used for \c{envir} ensures the correspondance between the textual
%% arguments of \c{h} and the arguments of \c{h2}.
%% \begin{lstlisting}
%%  sh <- substitute(h)
%%  fcall <- paste(sh, "(x, y)")
%%  ...
%%  h2 <- function(x, y)
%%  eval(parse(text = fcall),
%%  envir = list(x = x, y = y),
%%  enclos = parent.frame(2))
%% \end{lstlisting}

@@

\begin{compactitem}[---]
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote}, \c{expression} functions. Function arguments are passed as
  promises, and \c{substitute} is used to retrieve the source expression
  associated with the promise.
  \item {\it Reflection:} This group corresponds to uses of the \c{match.call}
  function to reflectively capture the expression that invoked the current
  function.
\item {\it String:} Finally, expressions can be created from strings by invoking
   \c{parse}, \c{str2expression}, or \c{str2lang}.
\end{compactitem}

\noindent
Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
This data is obtained by dynamically tainting values as they are produced by the
various sources. Due to technical reasons, there is some imprecision in the
results. In particular, we are not able to classify all sites. Manual inspection
of numerous examples suggests that the classified results are accurate.

@@

Strings could correlate with dynamic code loading. This is what the base
functions \c{source} and \c{sys.source} do. We observed few calls
(\packageNbParseFromFileSites in total) that consume the result of calling
\c{parse} on a file. Most of the calls build strings programmatically. We also
identified one function \c{invokeRestartInteractively} that prompts the user for
input, parses it, and passes it to \eval. The use of strings seems to correlate
with less sophisticated programmers; in our Kaggle corpus,
\kaggleParseExprsSitePercent of sites use strings.

\paragraph{Discussion}

% Compare to JS ? Everything comes from a string in JS anyway
% Composite (concatenation of strings), DOM and constant are the most common
% ...
The origin data suggests that constructing expressions from strings is a
minority of the use cases. Instead, the constructed category shows that most \evals comes from code that was processed by the compiler, and may be
slightly modified by the programmer before invoking \eval. Both constructed and
reflection categories roughly correspond to meta-programming. Some of these use
cases could likely be replaced by macros if the R designers could be convinced
to overcome their distaste for those.

>>>>

Contrary to the Javascript \eval, which does not specify the environment in
which to evaluate, the second argument of \eval in R, \c{envir}, is dedicated to
that. This argument determines what is visible to the computation started by
\eval and the potential reach of its side-effects. Environments used by \eval
can be classified into the following four kinds:

@@

\noindent
As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in the
scope of a function with global as a distant second. This means that most
variable reads and most side-effects operate on local variables. But for which
function? From the point of view of the caller of \eval,
Table~\ref{tab:funoffset} identifies that function by its offset on the call
stack. Thus 0 is the direct caller, 1 is its parent, and so on. The data
suggests that in 81\% of cases, \eval accesses the caller -- this means the
variables of the function where \eval textually occurs are read and written to.
It is noteworthy that 1.5\% of sites evaluate three frames or above. This means
that, in general, modular reasoning is impossible in R. To understand what any
given code snippet may do requires fully understanding all the functions that
may be called, transitively, from that snippet. The actions of \eval happen at a
distance. Finally, note that any site may see several kinds, but in
\packageNbOneCategoryEnvirSitePercent of the cases they have a single kind. It
means that a compiler could predict the \emph{kind} of environment after the
first invocation.

@@

\noindent
Global \evals are likely split between intentional and accidental one. Direct
references to the top-level, using \c{globalenv()} or \c{.GlobalEnv}, are rare;
they occur in only \packageNbExplicitGlobalSites sites. Thus we suspect most
uses of global are accidental. They arise from properties of our corpus, most
scripts are top-level code snippets. Thus, global is often the caller of \eval
or close to it. The reason we make this point is that values stored in the
global environment are visible to all functions and are not reclaimed by the
garbage collector. So accidental uses may pollute that namespace.

Each synthetic environment has parent that is specified when calling
\c{new.env}. When \c{envir} is a list or a data frame, \eval uses its third
argument (\c{enclos}). The parent is used to search for variables not found in
its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
kinds for synthetic environments. Most of them are functions, then come global
and package.

\paragraph{Discussion}
The data presented in this section is what one could expect. For {\it
  functions}, they represent the main user case for \eval. Thus \eval is really
extending the behavior of that function by executing a dynamically selected
piece of code in the function's scope. Typically, variables are read (mostly),
written (less), but there are also some cases where new variables are injected
or existing variables are deleted. Usually, this happens in the current
function, but also in frames arbitrarily far up the call stack. One data point
we lack is how the target environment was obtained. The expected case is that
the body of a promise was modified and the resulting expression was evaluated in
the same environment the promise originated from. Likely less frequent are cases
where \eval is provided the results of programmatically selecting some call
stack. For {\it synthetics}, their relatively high frequency correlates to use
cases where one wants to evaluate an expression in either a restricted
environment or to use a data structure as an environment. For {\it global}, its
relatively high frequency is likely an artifact of how the code of our corpus is
is run. For {\it packages}, only \packageNbPackageNamespaceEnvSites sites have
this kind. This is probably for the best as mutating the bindings of
a loaded package is improper and R tries to make it difficult.

@@

\begin{compactitem}[---]
\item {\it Constructed:} Expressions can be constructed by invoking the
  \c{quote}, \c{enquote}, \c{expression} functions. Function arguments are
  passed as promises, and \c{substitute} is used to retrieve the source
  expression associated with the promise.
\item {\it Reflection:} This group corresponds to uses of \c{match.call} to
  reflectively capture the expression that invoked the current function.
\item {\it String:} Expressions created from strings by invoking \c{parse},
  \c{str2expression}, or \c{str2lang}.
\end{compactitem}

\vspace{1mm}\noindent Table~\ref{tab:provenance} summarizes the expression
provenance in our corpus. This data is obtained by dynamically tainting values
as they are produced by the various sources. Due to technical reasons, there is
some imprecision in the results. In particular, we are not able to classify all
sites. Manual inspection of numerous examples suggests that errors are rare.

@@

Strings could correlate with dynamic code loading. This is what the base
functions \c{source} and \c{sys.source} do. We observed few calls
(\packageNbParseFromFileSites in total) that consume the result of calling
\c{parse} on a file. Most of the calls build strings programmatically. We also
identified one function \c{invokeRestartInteractively} that prompts the user for
input, parses it, and passes it to \eval. The use of strings seems to correlate
with less sophisticated programmers; in Kaggle,
\kaggleParseExprsSitePercent of sites use strings.

\paragraph{Discussion}
The origin data suggests that constructing expressions from strings is a
minority of the use cases. Instead, the constructed category shows that most
\evals comes from code that was processed by the compiler, and may be slightly
modified by the programmer before invoking \eval. Both constructed and
reflection categories roughly correspond to meta-programming. Some of these use
cases could likely be replaced by macros if the R designers could be convinced
to overcome their distaste for those.
